Staged typed type-preserving tagless ``interpreters'', \emph{finally}

### 2007-03-30 tagless/tagless-final.tex
Starting to work

\documentclass[preprint]{sigplanconf}
\usepackage{amsmath,amssymb}
\usepackage{natbib1}
\usepackage{comment}
\usepackage{url}

@@

\usepackage{natbib1}
\bibpunct();A{},
\let\cite=\citep

@@

@@

See Section 1 of Walid02. Our subject area is that of a definitional DSL
interpreter. 
Define Staging, meta-language and object language. (Walid uses
`subject' language).
\jacques{\emph{subject} and \emph{object} are both used in the current
write-up; should we define them to be equivalent, or just stick to one?}

@@

\oleg{should say somewhere in the Introduction that we are interested
  in the typed subject language embedded in the typed
  meta-language. Refer to Sec 2 (see the remark below) why this is
  good. We assume the DSL setting: we enter EDSL terms in the
  meta-language program (e.g., GHCi or (Meta)OCaml top-level). A
  different approach is reading EDSL terms from a file. In that case, we
  need to typecheck them first. We do not deal with that problem here,
  because it has been solved. First we mention Walid02 where they use
  dependent typed. However, these don't seem to be needed: refer to
  `typing dependent typing', and the related work section.}
\jacques{I think it is important to emphasize, probably later in the 
paper, that we could write a camlp4 extension (or simply a parser in
Haskell) that would transform a nice syntax for a lambda calculus into
syntactic terms for what we present here.  Even the self-interpreter
terms could be done this way, which would simplify things tremendously.}
\jacques{I quite agree that the point that needs to be revisited is
  that a \emph{direct} interpreter for a (typed) language "cannot" be
  done. see discussion in related work about the direct interpreter,
  with regards to Walid's paper. Perhaps bring some of it here?}
\jacques{I have looked at a fair bit of the literature, and I have not
  seen this done, especially not in the typed 'online' setting.  The
  one question that seems to be open is, how accurate can we make the
  PE without a tough BTA?}
Should we mention some of these points in our contributions subsection
below?

@@

Consider a language with abstraction, application and boolean
literals with de Bruin indices as variables. This is the language used
in the body of \cite{WalidIFCP02}, from which we borrow the first part of 
this example.

@@

Note that the subject language is typed. If we make a mistake
(for example, replace |int 7| with |bool True| in |testpowfix7|, the
compiler will highlight that term and complain that it couldn't match
the expected |Int| against the inferred |Bool|. A type error
also results when entering |lam (\x -> app x x)|, with the compiler
complaining about an infinite type. The type-checker of the
meta-language will flag meaningless subject level terms --
before any of their evaluation. If we forget, for example, |app| in
some of the terms above, we get an error message with a precise
location. So the type-checker also flags syntactically invalid
subject-level terms.\oleg{Should we perhaps move this up, in the
  intro? To justify why typed-metalanguages and the typed DSL are
  good. Or keep the explanation here but refer to it from the Introduction.}

@@

\bibliographystyle{plainnat}
\bibliography{tagless}
\end{document}

>>>>

\documentclass[preprint]{sigplanconf}
\usepackage{amsmath,amssymb}
\usepackage{comment}
\usepackage{hyphenat}
\usepackage{url}
\usepackage{prooftree1}

@@

\usepackage{natbib1}
\let\cite=\citep

@@

\newcommand{\fun}[1]{\mathopen{\lambda\mathord{#1}.\,}}
\newcommand{\fix}[1]{\mathopen{\mathrm{fix\,}\mathord{#1}.\,}}
\newcommand{\Forall}[1]{\mathopen{\forall\mathord{#1}.\,}}
\newcommand{\Exists}[1]{\mathopen{\exists\mathord{#1}.\,}}
\newcommand{\cond}[3]{\mathrm{if\,}#1\mathrm{\,then\,}#2\mathrm{\,else\,}#3}

% Make space around section headings no less than space around theorems
\makeatletter
\@tempdima\topsep \advance\@tempdima\parsep
\ifdim\@sectionbelowskip   <\topsep \@sectionbelowskip   \topsep \fi
\ifdim\@subsectionbelowskip<\topsep \@subsectionbelowskip\topsep \fi
\makeatother

% Match the rule inserted by \@makecaption in sigplanconf.cls
\setproofrulebreadth .33pt
\newenvironment{floatrule}
    {\hrule width \hsize height .33pt \vspace{.5pc}}
    {\par\addvspace{1ex}}

% Make some Rel symbols into Bin symbols instead
\DeclareMathSymbol{\to}{\mathbin}{symbols}{"21}
\DeclareMathSymbol{:}{\mathbin}{operators}{"3A}

@@

A popular way to define and implement a programming language is to embed it in
another \citep{reynolds-definitional}.  The embedded language is called the
\emph{object} or \emph{subject} language, whereas the embedding language is the
\emph{metalanguage}.  Embedding is especially appropriate for domain\hyp
specific object languages because it supports rapid prototyping and integration
with the host environment.  If the metalanguage supports \emph{staging}, then
the embedding can compile object programs to the metalanguage and avoid the
overhead of interpreting them on the fly \citep{WalidICFP02}.  A staged
definitional interpreter is thus a promising way to build a domain\hyp specific
language.

\begin{figure}
    \begin{floatrule}
    \begin{proofrules}
        \[ \[ [x:t_1] \proofoverdots e:t_2 \] \justifies \fun{x}e:t_1\to t_2 \]
        \[ \[ [f:t_1\to t_2] \proofoverdots e:t_1\to t_2 \] \justifies \fix{f}e:t_1\to t_2 \]
        \[ e_1:t_1\to t_2 \quad e_2:t_1 \justifies e_1 e_2: t_2 \]
        \[ \text{$n$ is an integer} \justifies n:\ZZ \]
        \[ \text{$b$ is a Boolean} \justifies b:\BB \]
        \[ e_1:\ZZ \quad e_2:\ZZ \justifies e_1+e_2:\ZZ \]
        \[ e_1:\ZZ \quad e_2:\ZZ \justifies e_1 \times e_2:\ZZ \]
        \[ e_1:\ZZ \quad e_2:\ZZ \justifies e_1 \le e_2:\BB \]
        \[ e:\BB \quad e_1:t \quad e_2:t \justifies \cond{e}{e_1}{e_2}:t \]
    \end{proofrules}
    \end{floatrule}
    \caption{Our typed object language}
    \label{fig:object}
\end{figure}

We focus on embedding a typed object language into a typed metalanguage.
\ccshan{Collect reasons to type here.}

@@

\oleg{should say somewhere in the Introduction that we are interested
  in the typed object language embedded in the typed
  meta-language. Refer to Sec 2 (see the remark below) why this is
  good. We assume the DSL setting: we enter EDSL terms in the
  meta-language program (e.g., GHCi or (Meta)OCaml top-level). A
  different approach is reading EDSL terms from a file. In that case, we
  need to typecheck them first. We do not deal with that problem here,
  because it has been solved. First we mention Walid02 where they use
  dependent typed. However, these don't seem to be needed: refer to
  `typing dependent typing', and the related work section.}
\jacques{I think it is important to emphasize, probably later in the 
paper, that we could write a camlp4 extension (or simply a parser in
Haskell) that would transform a nice syntax for a lambda calculus into
syntactic terms for what we present here.  Even the self-interpreter
terms could be done this way, which would simplify things tremendously.}
\jacques{I quite agree that the point that needs to be revisited is
  that a \emph{direct} interpreter for a (typed) language "cannot" be
  done. see discussion in related work about the direct interpreter,
  with regards to Walid's paper. Perhaps bring some of it here?}
\jacques{I have looked at a fair bit of the literature, and I have not
  seen this done, especially not in the typed 'online' setting.  The
  one question that seems to be open is, how accurate can we make the
  PE without a tough BTA?}
Should we mention some of these points in our contributions subsection
below?

@@

Consider a language with abstraction, application and boolean
literals with de Bruin indices as variables. This is the language used
by \citet{WalidICFP02}, from which we borrow the first part of 
this example.

@@

Note that the object language is typed. If we make a mistake
(for example, replace |int 7| with |bool True| in |testpowfix7|, the
compiler will highlight that term and complain that it couldn't match
the expected |Int| against the inferred |Bool|. A type error
also results when entering |lam (\x -> app x x)|, with the compiler
complaining about an infinite type. The type-checker of the
meta-language will flag meaningless object level terms --
before any of their evaluation. If we forget, for example, |app| in
some of the terms above, we get an error message with a precise
location. So the type-checker also flags syntactically invalid
object-level terms.\oleg{Should we perhaps move this up, in the
  intro? To justify why typed-metalanguages and the typed DSL are
  good. Or keep the explanation here but refer to it from the Introduction.}

@@

\bibliographystyle{mcbride}
\bibliography{tagless}
\end{document}
